### 2021-02-27 paper/base_model_specification.tex
Updated derivations

\setstretch{1}
\begin{flushright} 
    \end{flushright}
    \begin{center} \textbf{Linkage Cluster Paper Draft}

@@

\hypertarget{related-work}{%
\subsection{Related Work}\label{related-work}}

Sadinle (2017) proposed a Bayesian method for bipartite record linkage
that provides a foundation for much of our proposed method. In order to
ensure one-to-one matching, his Gibbs sampler conditions on the linkage
status of all other records when making each linkage assignment. While
this elegantly ensures transitivity, the probability distribution of the
linkage vector \(\mathbf{Z}\) is somewhat complicated, and does not
easily allow for inclusion of additional structure among records. Our
method however, motivated by recent decouple/recouple methods, initially
relaxes the one-to-one matching assumption in order to incorporate
linkage information differing by subgroups.

There has been some work done in the literature on conducting record
linkage in situations where feild accuracy differs by subgroups of the
data, but these are hampered by technical issues that make them
difficult to use in practice. Larson (2005) proposed Bayesian
heirarchical record linkage model that explicitly modelled the
differences in parameters across blocks of the data. In relying on
blocking however, he assumes that the blocking variable is error-free,
when in practice such a procedure could result in many missed linkages.
In contrast, our proposed method induces high probabilities of matching
within a linkage cluster while still allowing the possibility to match
across clusters (when the clustering field is recorded with error. Hu et
al (2015) proposed a method that used a logistic regression submodel
within a broader record linkage procedure to account for ``expected
changes'' in the record of the underlying entity based on subgroup
membership. This method however specified for use in temporal settings
when the true attributes of the record change with time, and it is
unclear how such a method would be implemented when records are
collected syncronously with error, or when time information is
unavailable. Additionally, the method is semi-supervised, requiring
training data to implement the logistic regression, and is that not well
suited for record linkage in practice. In contrast, the proposed method
is fully unsupervised and is shown to not require extensive tuning of
prior distributions, allowing for ease of implementation in a variety of
settings.

@@

\underline{Sample $\lambda^{(s+1)}|\mathbf{Z}^{s}$} As a function of
\(\lambda\), the linkage structure parameter \(\mathbf{Z}\) is sequence
of successes and failures, and therefore
\(p(\mathbf{Z}|\lambda) = \mathcal{L}(\lambda|\mathbf{Z})\) determined
only by the number of dupliates
\(D = \sum_{i=1}^{n_B}\mathbf{1}_{z_j < n_A + 1}\) encoded by
\(\mathbf{Z}\). Thus we have

@@

\underline{Sample $\mathbf{Z}^{(s+1)}|\Gamma, \mathbf{m}^{(s+1)}, \mathbf{u}^{(s+1)}, \lambda^{(s+1)}$}
Because we sample \(Z_j\) independently of all other \(Z_{j'}\), we use
only the full conditional for an individual \(Z_j\). Let \(\Gamma_{.j}\)
denote the set of \(n_A\) comparison vectors with \(j \in B\), and note
that as a function of \(Z_j\), the likelihood
\(p(\Gamma_{.j}|Z_j, \mathbf{m}, \mathbf{u})\) is a discrete
distribution with probalities given by
\[p(\Gamma_{.j}|Z_j = z_j, \mathbf{m}, \mathbf{u}) = \frac{\prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}}{\sum_{z_j = 1}^{n_B} \prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}} = \frac{\prod_{i=1}^{n_A}\left(\frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}}\right)^{\mathbf{1}_{z_j = i}}}{\sum_{z_j = 1}^{n_B}\prod_{i=1}^{n_A}\left(\frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}}\right)^{\mathbf{1}_{z_j = i}}} \]
\[\propto 
\begin{cases} 
    w_{ij}  & z_j \leq n_A; \\
    1 &  z_j  = n_A + 1 \\
\end{cases}\]

where where
\(w_{ij} = \frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}} = \frac{P(\boldsymbol{\gamma_{ij}}|Z_j = i)}{P(\boldsymbol{\gamma_{ij}} |Z_j \neq i)}\).
The interested reader should note that these are precisely the
likelihood ratios used in the Fellegi-Sunter model to classify matches
and non-matches, and we therefore refer to \(w_{ij}\) as the
\emph{Fellegi Sunter weights}.

@@

\newpage

\hypertarget{bayes-estimate}{%
\subsection{Bayes Estimate}\label{bayes-estimate}}

Sadinle 2017

\hypertarget{decouplerecouple}{%
\subsection{Decouple/Recouple}\label{decouplerecouple}}

One difficulty in record linkage is conducting linkage while taking
appropriate dependencies into account. In the context of this paper in
which we assume that each file is recorded without duplicates, this
means our linkage procedure should respect one-to-one matching. The
posterior sampling of \(\mathbf{Z}\) outlined above resolves half of
this issue; since we only sample one potential match for record \(j\) in
each Gibbs iteration, it is impossible to match \(j\) to two records at
once. This in itself is an improvement over the classic Fellegi-Sunter
method that labels record pairs as matches or non matches independently
of all other pairs. However, our posterior sampling of \(\mathbf{Z}\)
does allow for the possibility that \(Z_j = i = Z_{j'}\) for some
\(j' \neq j\), which cleary violates the assumption that
\(\mathbf{X}_B\) is recorded without duplicates. Sadinle 2017 resolves
this issue by adapting the set of candidate matches for every \(Z_j\),
but we forgo this rigidity and instead ``decouple'' the linkage vector
\(\mathbf{Z}\) for the computational advantantages of sampling \(Z_j\)
independently. In our formulation, such illegal matches occasionally
occur within individual Gibbs iterations due to simple randomness, but
they do not occur consistently enough between the same records to
measurably influence the eventual Bayes estimate of the linkage
structure.

However, if we want to ensure the theoretic purity of each MCMC sample
of the linkage structure, we can ``recouple'' \(\mathbf{Z}\) by simply
resolving these illegal matchings. Explictly, define clusters
\(R(i) := \{j | Z_j = i\}\), and define all clusters such that
\(|R(i)| > 1\) as \emph{resolution clusters}. Within each cluster,
resample the candidate matches according to their Fellegi-Sunter
weights. We show this step to be computationally inexpensive and to be a
reasonable replication of more organic one-to-one matching. This
procedure would be most useful in cases where there are many ``near
matches'' across files, but in most cases, we find this this step to be
unnecessary.

\newpage

\hypertarget{linkage-cluster-model}{%
\subsection{Linkage cluster model}\label{linkage-cluster-model}}

The primary contribution of this paper is the establishment of linkage
clusters, which allow the reliability parameters \(\mathbf{m}^f\), the
discernment parameters \(\mathbf{u}^f\), and also the linkage
probability parameter \(\lambda\) to vary according to certain
attributes in the record pairs. We shall see that this set up upweights
linkage within clusters and downweights linkage across clusters, thus
using attribute information in a way not yet explored in the literture.

Linkage cluster assignment occurs at both at the level of the records
and of the record pairs. Assign each record in each file an appropriate
linkage cluster \(S_k \in \{S_1, \ldots, S_k\}\), denoting the
membership record \(j\) in the \(k^{th}\) cluster by \(j \in S_k\). We
assign the record pair comparison vector \(\gamma_{ij}\) to a cluster
only when both records belong to that cluster; otherwise we assign it to
\(S_0\). For example
\[i \in S_1, j \in S_1 \implies \gamma_{ij} \in S_1\]
\[i \in S_1, j \in S_2 \implies \gamma_{ij} \in S_0\] Cluster
assignments at the record pair level are necessary for the updating the
\(\mathbf{m}\) and \(\mathbf{u}\) parameters. We denote the reliability
and discernment parameters assoicated with each field for each cluster
as \(\mathbf{m}^{fk}\) and \(\mathbf{u}^{fk}\) respectively. To improve
readability, we modify the definition of \(\mathbf{m}_{ij}\) to
incorporate the linkage cluster assignment. Explicitly,
\[\mathbf{m}_{ij} = \prod_{f=1}^{F}\prod_{l=1}^{L_f}\left[ \left(m^{flk}\right)^{\mathbf{1}_{\gamma_{ij} \in S_k}}\left(m^{fl0}\right)^{\mathbf{1}_{\gamma_{ij} \in S_0}}\right]^{\mathbf{1}_{\gamma_{ij}^f = l}}\]
with an analogous definition for \(\mathbf{u}_{ij}\).

Additionally, define the vector of linkage probabilities
\(\Lambda_k = (\lambda_{k1}, \lambda_{k2}, \lambda_{k3})\), where
\(\lambda_{k1} = P(z_{j\in S_k} \in S_k)\) is the probability that a
record in \(\mathbf{X}_{B_k}\) has a match within its cluster in
\(\mathbf{X}_{A_k}\), \(\lambda_{k2} = P(z_{j \in S_k} \in S_{k'})\) is
the probability that such a record \(\mathbf{X}_{B_k}\) has a match
outside of its cluster \(\mathbf{X}_{A_{k'}}\) for some \(k' \neq k\),
and \(\lambda_{k3} = P(z_{j\in S_k} = n_A + 1)\) is the probability of a
non-match. This is a multinomial distribution where we necessarily have
\(\lambda_{k3} = 1 - \lambda_{k1} - \lambda_{k2}\). These parameters use
linkage cluster assignment at the record level.

Our new model specification becomes

\[P(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}, \boldsymbol{\theta}) =\prod_{k=1}^K \prod_{j \in \mathbf{X}_B^k}  \left[\lambda_{k1}^{\mathbf{1}_{z_j \in S_k}}\lambda_{k2}^{\mathbf{1}_{z_j \in S_{k'}}} \lambda_{k3}^{\mathbf{1}_{z_j = n_A + 1}}\prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}\right]\]

\[\mathbf{m^{fk}} \sim \text{Dirichlet}(\alpha^{f1k}, \ldots, \alpha^{fL_fk})\]
\[\mathbf{u^{fk}} \sim \text{Dirichlet}(\beta^{f1k}, \ldots, \beta^{fL_fk})\]
\[Z_{j \in S_k} | \Lambda_k =
\begin{cases} 
    \frac{1}{n_{A_k}}\lambda_{k1}  & z_j \in S_k; \\
    \frac{1}{n_A - n_{A_k}}\lambda_{k2}  & z_j \in S_{k'}; \\
     \lambda_{k3} & z_j = n_A + 1 \\
\end{cases}\]

\[\Lambda \sim \text{Dirichlet}(\Lambda_0)\] where
\(\Lambda_0 = (\Lambda_{k1}, \Lambda_{k2}, \Lambda_{k3}).\)

\textbf{THIS WOULD GO IN APPENDIX}: Through similar derivations, this
yields full conditionals: The \(\mathbf{m}\) and \(\mathbf{u}\)
parameters are updated through standard multinomial-dirichlet mechanics.
Thus we have
\[\mathbf{m^{fk}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\alpha^{f1k}(\mathbf{Z}), \ldots, \alpha^{fL_fk}(\mathbf{Z}))\]

\[\mathbf{u^{fk}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\beta^{f1k}(\mathbf{Z}), \ldots, \beta^{fL_fk}(\mathbf{Z}))\]
where
\[\alpha_{flk}(\mathbf{Z})= \sum_{i,j, k} \mathbf{1}_{obs(\gamma_{ij}^f)}\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j = i}\mathbf{1}_{\gamma_{ij} \in S_k}\]

\[\text{and } \beta_{flk}(\mathbf{Z})= \sum_{i,j, k} \mathbf{1}_{obs(\gamma_{ij}^f)}\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j \neq i}\mathbf{1}_{\gamma_{ij}^f = l}\]

We also have straightforward posterior distritutions for the linkage
probability vector \(\Lambda_k\):

\[\Lambda_k | \mathbf{Z} \sim \text{Dirichlet}(\mathbf{D_k} + \Lambda_0)\]

where \(\mathbf{D}_k = (D_{k1}, D_{k2}, D_{k3})\) with
\[D_{k1} = \sum_{j \in \mathbf{X}_{B_k}} \mathbf{1}_{z_j \in S_k}\]
\[D_{k2} = \sum_{j \in \mathbf{X}_{B_k}} \mathbf{1}_{z_j \in S_{k'}}\]
\[\text{and } D_{k3} = \sum_{j \in \mathbf{X}_{B_k}} \mathbf{1}_{z_j = n_A + 1}\]

Our full conditional for \(Z_j\) follows a familiar form:

\[Z_{j \in S_k}^{(s+1)} | \mathbf{m}^k, \mathbf{u}^k, \Gamma, \Lambda_k =
\begin{cases} 
    \frac{\lambda_{k1}}{n_{A_k}} w_{ij} & z_j \in S_k; \\
    \frac{\lambda_{k2}}{n_A - n_{A_k}} w_{ij}  & z_j \in S_{k'}; \\
     \lambda_{k3} & z_j = n_A + 1 \\
\end{cases}\]

And after integrating over \(\Lambda_k\) and rearraning terms,

\[Z_{j \in S_k}^{(s+1)} | \mathbf{m}^k, \mathbf{u}^k, \Gamma,  =
\begin{cases} 
    \frac{D_{k1} + \Lambda_{01}}{n_{A_k}(n_B - D_{k1} - D_{k2} + \Lambda_{03})} w_{ij} & z_j \in S_k; \\
    \frac{D_{k2} + \Lambda_{02}}{(n_A - n_{A_k})(n_B - D_{k1} - D_{k2} + \Lambda_{03})} w_{ij}  & z_j \in S_{k'}; \\
     1 & z_j = n_A + 1 \\
\end{cases}\]

\newpage

Observe two reasonable choices for \(p(Z, \lambda)\):

If \(\lambda \sim \text{Beta}(1, \frac{1}{n_A})\), then the prior for
\(Z\), integrated over \(\lambda\), becomes

\[=\begin{cases} 
    1  & z_j \leq n_A; \\
    1 &  z_j  = n_A + 1 \\
\end{cases}\] Thus this is a uniform prior on the labellings of \(Z\). A
second useful prior would be \(\lambda \sim \text{Beta}(1, 1)\), which
yields \[=\begin{cases} 
    1  & z_j \leq n_A; \\
    n_A &  z_j  = n_A + 1 \\
\end{cases}\] Here, a match and a non-match are equaly likely, and among
the matches, all records are likely to be a match. There is little
difference between the the two priors in practice, as the data should
overpower the prior.

The \(\mathbf{m}\) and \(\mathbf{u}\) parameters are updated through
standard multinomial-dirichlet mechanics. Thus we have

Since \(\lambda\) provides the probability that a record in
\(\mathbf{X}_B\) has match in \(\mathbf{X}_A\), it has a full
conditional distribution

\[\lambda|\mathbf{Z^{(s)}} \sim \text{Beta}(D + \alpha_{\lambda}, n_B - D + \beta_{\lambda})\]
where \(D = \sum_1^{n_B} (Z_j < n_A + 1)\) is the number of records
identified as matches in the previous Gibbs iteration.

Lastly, we sample each \(Z_j\) according to these posterior
distributions. To do so, note that as a function of \(Z_j\), the
likelihood is a discrete distibution with probabilities proportional to

\[p(\Gamma_{.j}|Z_{j}) \propto \mathbf{1}_{z_j < n_A + 1}\lambda\prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}} + \mathbf{1}_{z_j = n_A + 1}(1-\lambda)\prod_{i=1}^{n_A}\mathbf{u}_{ij}\]
\[\propto \mathbf{1}_{z_j < n_A + 1}\lambda\prod_{i=1}^{n_A}\left(\frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}}\right)^{\mathbf{1}_{z_j = i}}+ \mathbf{1}_{z_j = n_A + 1}(1-\lambda)\]

\[\propto \lambda \left[\sum_{i=1}^{n_A}w_{ij}\mathbf{1}_{z_j = i}\right] + (1 - \lambda)\mathbf{1}_{z_j = n_A + 1}\]
where
\(w_{ij} = \frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}} = \frac{P(\boldsymbol{\gamma_{ij}}|Z_j = i)}{P(\boldsymbol{\gamma_{ij}} |Z_j \neq i)}\).
The interested reader should note that these are precisely the
likelihood ratios used in the Fellegi-Sunter model to classify matches
and non-matches, and we therefore refer to \(w_{ij}\) as the
\emph{Fellegi Sunter weights}.

The full conditional for \(Z_j\) is thus given by
\[p(Z_j^{(s+1)}|\mathbf{m}, \mathbf{u}, \Gamma, \lambda, \mathbf{Z^{(s)}}) \propto p(\Gamma_{.j}|\mathbf{m}, \mathbf{u}, Z_j, \lambda)\overbrace{p(\mathbf{m}, \mathbf{u}|\mathbf{Z^{(s)}})}^{\text{proportionality}}p(Z_j | \lambda) \overbrace{p(\lambda|\mathbf{Z^{(s)}})}^{\text{proportionality}}\]
\[\propto \left(\sum_{i=1}^{n_A}w_{ij}\mathbf{1}_{z_j = i} + \mathbf{1}_{z_j = n_A + 1}\right)\left(\lambda\sum_{i=1}^{n_A}\frac{1}{n_A}\mathbf{1}_{z_j = i} + (1-\lambda)\mathbf{1}_{z_j = n_A + 1}\right)\]
\[= \frac{\lambda}{n_A}\sum_{i=1}^{n_A}w_{ij}\mathbf{1}_{z_j = i} + (1-\lambda)\mathbf{1}_{z_j = n_A + 1} \]
\[ \implies Z_j^{(s+1)} | \mathbf{m}, \mathbf{u}, \Gamma, \lambda \propto
\begin{cases} 
    \frac{\lambda}{n_A}w_{ij}   & z_j \leq n_A; \\
     1-\lambda &  z_j  = n_A + 1 \\
\end{cases}\]

The relationship between this sampling mechanism and that of Sadinle
2017 warrants discussion.

>>>>

\setstretch{1}
\begin{flushright} 
    \end{flushright}
    \begin{center} \textbf{Base Model Specification}

@@

@@

\textbackslash underline\{Sample \(\lambda^{(s+1)}|\mathbf{Z}^{(s)}\) As
a function of \(\lambda\), the linkage structure parameter
\(\mathbf{Z}\) is sequence of successes (when \(z_j < n_A + 1\)) and
failures (when \(z_j = n_A + 1\)), and therefore
\(p(\mathbf{Z}|\lambda) = \mathcal{L}(\lambda|\mathbf{Z})\) is
determined only by the number of dupliates
\(D = \sum_{i=1}^{n_B}\mathbf{1}_{z_j < n_A + 1}\) encoded by
\(\mathbf{Z}\). Thus we have

@@

\underline{Sample $\mathbf{Z}^{(s+1)}|\Gamma, \mathbf{m}^{(s+1)}, \mathbf{u}^{(s+1)}, \lambda^{(s+1)}$}
Because we sample \(Z_j\) independently of all other \(Z_{j'}\), we use
only the full conditional for an individual \(Z_j\). Let \(\Gamma_{.j}\)
denote the set of \(n_A\) comparison vectors with \(j \in B\), and note
that as a function of \(Z_j\), the likelihood
\(p(\Gamma_{.j}|Z_j, \mathbf{m}, \mathbf{u})\) is a discrete
distribution with probalities proportional to
\[p(\Gamma_{.j}|Z_j = z_j, \mathbf{m}, \mathbf{u}) \propto \prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}\]
\[\propto \prod_{i=1}^{n_A}\left(\frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}}\right)^{\mathbf{1}_{z_j = i}} \;\;\;\;\;\;\;\; \text{Divide through by} \prod_{i = 1}^{n_A}\mathbf{u}_{ij}\]
\[=
\begin{cases} 
    w_{ij}  & z_j \leq n_A; \\
    1 &  z_j  = n_A + 1 \\
\end{cases}\]

where
\(w_{ij} = \frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}} = \frac{P(\boldsymbol{\gamma_{ij}}|Z_j = i)}{P(\boldsymbol{\gamma_{ij}} |Z_j \neq i)}\).
The interested reader should note that these are precisely the
likelihood ratios used in the Fellegi-Sunter model to classify matches
and non-matches, and we therefore refer to \(w_{ij}\) as the
\emph{Fellegi Sunter weights}.

@@
